Well-quasi-orderings on word languages

### 2025-10-16 src/amalgam.tex
fix: reworking conclusion

\section{Infixes and Amalgamation Systems}
\label{infixes-amalgamation:sec}

@@

\begin{theorem}
	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be an \kl{effective amalgamative class} of languages.
    Then the following are equivalent:
	\begin{enumerate}
        \item\label{wqo-infix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{infix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{wqo-prefix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{prefix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{emptiness-decidable} Emptiness is decidable for languages in $\mathcal{C}$.
	\end{enumerate}
\end{theorem}

%\AP We say that a \kl{strongly effective amalgamative class} is an
%\kl{effective amalgamative class} for which the emptiness problem is decidable.
%Let us immediately remark that the class of \kl{regular languages} is
%\kl(amalg){strongly effective}, and so is the class of context-free languages.
%Therefore, \cref{infix-wqo-is-emptiness:thm} provides a concrete decision
%procedure for these classes.

@@

If we additionally assume that such a language is closed under infixes, we
obtain an even stronger structure: All such languages are regular!

@@

\AP To construct a decision procedure for well-quasi-orderedness under
$\infleq$, we our \kl{amalgamation systems} to satisfy certain
\intro(amalgamation){effectiveness assumptions}. We require that for an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$, $R$ is recursively
enumerable, the function $\canrun(\cdot)$ is computable, and for any two runs
$\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable. Additionally, we
require the class to be effectively closed under rational transductions
\cite[Chapter 5, page 64]{BERST79}.

%\AP We say that a class $\mathcal{C}$ of languages is an \intro{effective
%amalgamative class} whenever for every $L \in \mathcal{C}$, there exists an
%effective \kl{amalgamation system} recognizing $L$, and such that $\mathcal{C}$
%is \kl{effectively closed under rational transductions}. Recall that a
%\intro{rational transduction} is a \emph{relation} $R \subseteq \Sigma^* \times
%\Gamma^*$ that can be defined by a (nondeterministic) finite automaton with
%output \cite[Chapter 5, page 64]{BERST79}. A class of languages $\mathcal{C}$
%is \intro{effectively closed under rational transductions} when, for every
%language $L \in \mathcal{C}$, and every rational transduction $R \subseteq
%\Sigma^* \times \Gamma^*$, the image of $L$ through $R$ --- that is, $\setof{v
%\in \Gamma^*}{\exists u \in L. (u,v) \in R}$ --- is in $\mathcal{C}$ and
%effectively computable. In particular, it implies that the class contains
%languages over arbitrary (finite) alphabets.

%\AP A class $\mathcal{C}$ of languages is \intro(amalg){strongly effective}
%whenever the emptiness problem for languages in $\mathcal{C}$ is decidable.
%This notion is interesting because usual language problems such as boundedness
%or simultaneous boundedness are decidable for \kl{strongly effective
%amalgamative classes}~\cite{ASZZ24}.

The class $\mathcal{C}_\text{aut}$ of \kl{regular languages} and the class
$\mathcal{C}_{\text{cfg}}$ of context-free languages are examples of
\kl{effective amalgamative classes}, hence the following corollary.

>>>>

% LTeX: language=en-GB 
% !TeX root=../wqo-on-words.lncs.tex
\section{Infixes and Amalgamation Systems}
\label{infixes-amalgamation:sec}

@@

\begin{theorem}
	\label{infix-wqo-is-emptiness:thm}
  \proofref{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be an \kl{effective amalgamative class} of languages.
    Then the following are equivalent:
	\begin{enumerate}
        \item\label{wqo-infix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{infix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{wqo-prefix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{prefix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{emptiness-decidable} Emptiness is decidable for languages in $\mathcal{C}$.
	\end{enumerate}
\end{theorem}

@@

If we additionally assume that such a language is closed under taking infixes,
we obtain an even stronger structure: All such languages are regular!

@@

\AP To construct a decision procedure for well-quasi-orderedness under
$\infleq$, we need our \kl{amalgamation systems} to satisfy certain
\intro(amalgamation){effectiveness assumptions}. We require that for an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$, $R$ is recursively
enumerable, the function $\canrun(\cdot)$ is computable, and for any two runs
$\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable. Additionally, we
require the class to be effectively closed under rational transductions
\cite[Chapter 5, page 64]{BERST79}.

Under these assumptions, one can transform the inclusion test of
\cref{bounded-language:eq} of \cref{bounded-language:thm} into an effective
procedure, using pumping arguments from \cite[Section 4.2]{ASZZ24}, which, in
turn, allows us to prove \cref{infix-wqo-is-emptiness:thm}. Since the class
$\mathcal{C}_\text{aut}$ of \kl{regular languages} and the class
$\mathcal{C}_{\text{cfg}}$ of context-free languages are examples of
\kl{effective amalgamative classes}, the following corollary is immediate.

### 2025-10-16 src/conclusion.tex
fix: trying my best
fix: reworking conclusion

We provided concretes statements that justify why the \kl{subword relation}
is used when defining \kl{well-quasi-orders} on finite words. Even if
\kl{prefix}, \kl{suffix} or \kl{infix} relations are meaningful, they are
\kl{well-quasi-ordered} if and only if they behave similarly to disjoint copies
of $\Nat$ or $\Nat^2$. However, our approach suffers some limitations 
and opens the road to natural continuation of this line of work.

\paragraph*{Towards infinite alphabets} In this paper, we restricted our
attention to \emph{finite} alphabets, having in mind the application to
\kl{regular languages}. However, the conclusions of
\cref{bounded-language:thm}, \cref{small-ordinal-invariants:thm}, and
\cref{prefixes:thm} could be conjecture to hold in the case of infinite
alphabets (themselves equipped with a \kl{well-quasi-ordering}). This would
require new techniques, as the finiteness of the alphabet is crucial to all of
our positive results.

\paragraph*{Monoid equations}  It could be interesting to understand which
monoids $M$ recognize languages that are \kl{well-quasi-ordered} by the
\kl{infix}, \kl{prefix} or \kl{suffix} relations. This research direction is
connected to finding which classes of graphs of \emph{bounded clique-width} are
\kl{well-quasi-ordered} with respect to the \emph{induced subgraph relation},
as shown in \cite{DRT10}, and recently revisited by one of the authors in
\cite{L24:arxiv:v2}.

\paragraph*{Complexity} We have chosen to disregard complexity considerations
when proving decidability results, as we do not believe that a fined grained
complexity analysis would be particularly enlightening. However, we conjecture
that for regular languages, the complexity should be polynomial in the size of
the minimal automaton recognizing the language.

@@

\paragraph*{Graph Classes}
\todo[inline]{aliaume: retalk about \cite{ALM17} and \cite{L24:arxiv:v2}}

>>>>

We have described the landscapes of \kl{well-quasi-ordered} languages for the
natural orderings on finite words: \kl{prefix}, \kl{suffix}, and \kl{infix}
relations. While the prefix and suffix relation exhibit very simple behaviours,
the infix relation can encode many complex quasi-orders (and even simulate the
\kl{subword ordering}). In the case of languages that are described by simple
computational models, or languages that are ``structurally simple''
(\kl{bounded languages}, \kl{downwards closed} languages), we showed that only
very simple \kl{well-quasi-orders} can be obtained: they are essentially
isomorphic to disjoint unions of copies of finite sets, $(\Nat, \leq)$, and
$(\Nat^2, \leq)$. Finally, under effectiveness assumptions on the language
(such as being recognized by an \kl{amalgamation system}, or being the set of
infixes of an \kl{automatic sequence}), we proved the decidability of being
\kl{well-quasi-ordered} for the \kl{infix relation}. We believe that these very
encouraging results pave the way for further research on deciding which sets
are \kl{well-quasi-ordered} for other orderings. Let us now discuss
some possible research directions and remarks.

\paragraph*{Towards infinite alphabets} In this paper, we restricted our
attention to \emph{finite} alphabets, having in mind the application to
\kl{regular languages}. However, the conclusions of
\cref{bounded-language:thm}, \cref{small-ordinal-invariants:thm}, and
\cref{prefixes:thm} could be conjectured to hold in the case of infinite
alphabets (themselves equipped with a \kl{well-quasi-ordering}). This would
require new techniques, as the finiteness of the alphabet is crucial to all of
our positive results.

\paragraph*{Monoid equations}  It could be interesting to understand which
monoids $M$ recognize languages that are \kl{well-quasi-ordered} by the
\kl{infix}, \kl{prefix} or \kl{suffix} relations. This research direction is
connected to finding which classes of graphs of \emph{bounded clique-width} are
\kl{well-quasi-ordered} with respect to the \emph{induced subgraph relation},
as shown in \cite{DRT10}, and recently revisited in \cite{lopez25}.

@@

\paragraph*{Factor Complexity} \AP Let us conclude this section with a few
remarks on the notion of \kl{factor complexity} of languages. Recall that the
\intro{factor complexity} of a language $L \subseteq \Sigma^*$ is the function
$f_L: \Nat \to \Nat$ such that $f_L(n)$ is the number of distinct words of size
$n$ in $L$. We extend the notion of \kl{factor complexity} to finite, infinite,
and bi-infinite words as the \kl{factor complexity} of their set of finite
\kl{infixes}. For the \kl{prefix relation} and the \kl{suffix relation}, all
\kl{well-quasi-ordered} languages have a bounded \kl{factor complexity}, since
they are finite unions of \kl{chains}.

While there clearly are languages with low \kl{factor complexity} that are not
\kl{well-quasi-ordered} for the \kl{infix relation}, such as the language $L
\defined \dwset{ a b^* a }$; one would expect that languages that are
\kl{well-quasi-ordered} for the \kl{infix relation} would have a low \kl{factor
complexity}.

In some sense, our results confirm this intuition in the case of languages
described by a simple computational model. For languages recognized by
\kl{amalgamation systems}, being \kl{well-quasi-ordered} implies being a
\kl{bounded language}, and therefore being  included in some finite union of
languages of the form $w_1^* w_2 w_3^*$. Hence, these languages have at most a
quadratic \kl{factor complexity}. This is also the case for languages described
as the infixes of a finite set of pairs of \kl{morphic sequences}. Indeed, the
factor complexity of a \kl{morphic sequence} that is \kl{uniformly recurrent}
is linear \cite[Theorem 24]{NIPR09}, therefore the \kl{factor complexity} of a
language given by \kl{sequence representation} using \kl{morphic sequences} is
at most quadratic.

However, there are \kl{downwards closed} languages that are
\kl{well-quasi-ordered} for the \kl{infix relation} but have an \kl{exponential
factor complexity}: the $(5, 3)$-Toeplitz word is \kl{uniformly recurrent}
\cite[p. 499]{CAKA97}, but has an exponential factor complexity \cite[Theorem
5]{CAKA97}. This shows that our computational models somehow fail to capture
vast classes of \kl{well-quasi-ordered} languages with a high \kl{factor
complexity}. It would be interesting to understand which new proof techniques
would be required to obtain decidability for these languages.

### 2025-10-16 src/introduction.tex
fix: reworking conclusion

\AP Let us argue that these decision procedures fit a larger picture in the
research area of well-quasi-orderings.
Indeed, there has been recent breakthroughs in deciding whether a given order
is a \kl{well-quasi-order}, for instance in the context of the verification of
infinite state transition systems~\cite{DBLP:conf/fsttcs/FinkelG19} or in the
context of logic~\cite{DBLP:journals/pacmpl/BergstrasserGLZ24}.
In the graph theory community, recent works have studied classes of graphs 
that are \kl{well-quasi-ordered} by the \kl{induced subgraph relation}
using similar language theoretic techniques~\cite{DRT10,L24:arxiv:v2,ALM17}.
Furthermore, a previous work by Kuske shows that any
\emph{reasonable}\footnote{ This will be made precise in
\cref{infix-embedding:thm}. } partially ordered set $(X, \leq)$ can
be embedded into $\set{a,b}^*$ with the \kl{infix relation}~\cite[Lemma
5.1]{DBLP:journals/ita/Kuske06}. Phrased differently, one can encode a large
class of partially ordered sets as subsets of $\set{a,b}^*$. As a consequence,
the following decision problem provides a reasonable abstract framework for
deciding whether a given partially ordered set is \kl{well-quasi-ordered}:
given a language $L \subseteq \Sigma^*$, decide whether $L$ is
\kl{well-quasi-ordered} by the \kl{infix relation}.

>>>>

\AP Let us argue that these decision procedures fit a larger picture in the
research area of well-quasi-orderings.
Indeed, there has been recent breakthroughs in deciding whether a given order
is a \kl{well-quasi-order}, for instance in the context of the verification of
infinite state transition systems~\cite{DBLP:conf/fsttcs/FinkelG19} or in the
context of logic~\cite{DBLP:journals/pacmpl/BergstrasserGLZ24}.
In the graph theory community, recent works have studied classes of graphs 
that are \kl{well-quasi-ordered} by the \kl{induced subgraph relation}
using similar language theoretic techniques~\cite{DRT10,lopez25,ALM17}.
Furthermore, a previous work by Kuske shows that any
\emph{reasonable}\footnote{ This will be made precise in
\cref{infix-embedding:thm}. } partially ordered set $(X, \leq)$ can
be embedded into $\set{a,b}^*$ with the \kl{infix relation}~\cite[Lemma
5.1]{DBLP:journals/ita/Kuske06}. Phrased differently, one can encode a large
class of partially ordered sets as subsets of $\set{a,b}^*$. As a consequence,
the following decision problem provides a reasonable abstract framework for
deciding whether a given partially ordered set is \kl{well-quasi-ordered}:
given a language $L \subseteq \Sigma^*$, decide whether $L$ is
\kl{well-quasi-ordered} by the \kl{infix relation}.
